### 2023-05-30 Manuscript/Draft1.Rmd
manuscript update

We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings). VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online tutorial https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/. After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed results and suggested we also consider the options in Settings A. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.

@@

@@

 Of the ~27% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

>>>>

We follow the advice given in [@thorson2019guidance] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. The covariates we considered were the dynamic temperature values and static habitat values ($Hab_{J.s}$) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings). VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal]. After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A.

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.

@@

To test the impact of shifting populations on abundance estimates, we simulated known population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the true values. We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation. We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included. Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.

Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend. While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error. This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses. This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs. Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings **cite Chris C paper**. The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates. Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios. While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances. For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.

something about seasonal impact seen in all estimates with fall estimates typically noticeably worse. We explored this through the ratio plots. This was mostly a problem with VAST as SM showed more consistent estimates between seasons.

@@

 Of the ~27% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

### 2023-05-30 Manuscript/Draft1.tex
manuscript update

We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online tutorial \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}. After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed results and suggested we also consider the options in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct). The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.

@@

@@

Of the \textasciitilde27\% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

>>>>

We follow the advice given in (Thorson 2019) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. The covariates we considered were the dynamic temperature values and static habitat values (\(Hab_{J.s}\)) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.). After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A.

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.

@@

To test the impact of shifting populations on abundance estimates, we simulated known population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the true values. We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation. We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included. Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.

Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend. While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error. This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses. This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs. Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings \textbf{cite Chris C paper}. The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates. Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios. While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances. For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.

something about seasonal impact seen in all estimates with fall estimates typically noticeably worse. We explored this through the ratio plots. This was mostly a problem with VAST as SM showed more consistent estimates between seasons.

@@

Of the \textasciitilde27\% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ThorsonSeasonal}{}}%
Thorson, James T. (n.d.). {Seasonal and Interannual Dynamics Github Documentation}. {https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
